Laplacian envelope

### 2020-06-22 Notes/laplace.tex
graph of optimised sticky models
started alterative optimisation approach

% -*- TeX -*- -*- UK -*-
% ----------------------------------------------------------------
% arXiv Paper ************************************************
%
% Subhaneil Lahiri's template
%
% Before submitting:
%    Comment out hyperref
%    Comment out showkeys
%    Replace \input{?.tex} with its contents
%       or include ?.tex in zip/tar file
%    Put this file, the .bbl file, any picture or
%       other additional files and natbib.sty
%       file in a zip/tar file
%
% **** -----------------------------------------------------------
\documentclass[12pt]{article}
%Preamble:
\usepackage{a4wide}
\input{sl_preamble.tex}
%
% >> Only for drafts! <<
\usepackage[notref,notcite]{showkeys}
%
\input{sl_graphics_preamble.tex}
\graphicspath{{Figures/}}
%
% ----------------------------------------------------------------
%\numberwithin{equation}{section}
%\renewcommand{\baselinestretch}{1.5}
% ----------------------------------------------------------------
% New commands etc.
\input{sl_definitions.tex}
\input{sl_symbols.tex}
%matrices
%vec of ones
\newcommand{\onev}{\mathbf{e}}
%mat of ones
\newcommand{\onem}{\mathbf{E}}
%Markov matrix
\newcommand{\MM}{\mathbf{Q}}
%prob distribution
\newcommand{\pr}{\mathbf{p}}
%equilibrium distribution
\newcommand{\eq}{\pr^\infty}
%first passage times
\newcommand{\fpt}{\mathbf{T}}
%off-diag first passage times
\newcommand{\fptb}{\overline{\fpt}}
%fundamental matrix
\newcommand{\fund}{\mathbf{Z}}
%other symbols
\newcommand{\Pb}{\mathbf{P}}
\newcommand{\D}{\mathbf{D}}
\newcommand{\pib}{\boldsymbol{\pi}}
\newcommand{\Lb}{\boldsymbol{\Lambda}}
\newcommand{\etw}{\boldsymbol{\eta}^w}
\newcommand{\thb}{\boldsymbol{\theta}}
%synapse models
\newcommand{\w}{\mathbf{w}}
\newcommand{\W}{\mathbf{W}}
\newcommand{\M}{\mathbf{M}}
\newcommand{\enc}{\mathbf{K}}
\newcommand{\frg}{\W^{\mathrm{F}}}
\newcommand{\F}{\boldsymbol{\Phi}}
%superscripts
\newcommand{\pot}{^{\text{pot}}}
\newcommand{\dep}{^{\text{dep}}}
\newcommand{\potdep}{^{\text{pot/dep}}}
%with activity indep processes
\renewcommand{\hom}{\mathbf{H}}
\newcommand{\Mh}{\widetilde{\M}}
\newcommand{\frgh}{\widetilde{\W}^{\mathrm{F}}}
\newcommand{\ench}{\widetilde{\enc}}
%eigenvectors
\newcommand{\evr}{\mathbf{u}}
\newcommand{\evl}{\boldsymbol{\eta}}
%snr curves etc
\newcommand{\syn}{\vec{w}}
\newcommand{\synid}{\syn_\text{id}}
\DeclareMathOperator{\SNR}{SNR}
\DeclareMathOperator{\snr}{SNR}
\DeclareMathOperator{\NNR}{NNR}
\DeclareMathOperator{\nnr}{NNR}
\newcommand{\snrb}{\overline{\snr}}
\DeclareMathOperator{\env}{Env}
\newcommand{\rh}{\hat{r}}
\newcommand{\CA}{\mathcal{A}}
\newcommand{\CB}{\mathcal{B}}
%avoid double superscripts
\renewcommand{\pdiff}[2]{\frac{\partial #1}{\partial #2}}

@@

In reality, $M$ can only take even integer values.
However, one can find a heuristic approximation to the envelope by maximising \wrt $M$.
We find that
%
\begin{multline}\label{eq:unienv}
  M_* = \pm\frac{2y_*}{\beta},
  \quad
  s = 2\sinh^2\frac{y_*}{M_*},
  \qquad
  \hat{A}_*(s) = \prn{\frac{ 2\sinh^2 \frac{y_*}{2} }{ y_* \cosh y_* } } \frac{\abs{\beta}}{s},
  \\ \text{where}\quad
  y_* = \tanh \frac{y_*}{2} \cosh y_* \approx 1.50553.
  \qquad
\end{multline}
%
This is only valid for
%
\begin{equation}\label{eq:univalid}
  2 \leq M_* \leq M,
  \qquad\implies\qquad
  2\sinh^2 \frac{y_*}{M} \leq s \leq 2\sinh^2 \frac{y_*}{2}.
\end{equation}
%

When $M \gg 2$, the validity region is approximately
%
\begin{equation}\label{eq:univalidapprox}
  \frac{4.53327}{M^2} \lesssim s \lesssim 1.36423.
\end{equation}
%
When $s \sim \CO(M^{-2}) \ll 1$, we have \(\beta \approx \pm \sqrt{2s}\) and the heuristic envelope is approximately
%
\begin{equation}\label{eq:unienvapprox}
  \hat{A}_*(s) \approx \frac{ 0.54203 }{ \sqrt{s} }.
\end{equation}
%

@@

@@

We can maximise this \wrt $\varepsilon$.
This results in a complicated quadratic equation for $\varepsilon$ with an even worse solution that we will not reproduce here.
We can find the boundary of the region of applicability by seeing at what value of $s$ is the optimal value of $\varepsilon=0$:
%
\begin{equation}\label{eq:stickyvalid}
\begin{aligned}
  \left.\pdiff{\hat{A}(s)}{\varepsilon}\right|_{\varepsilon=0} &\propto
    m \prn{S(m\beta) - S((m-1)\beta)} + (m-1) S(m\beta) \prn{S(m\beta) + 1} = 0,\\
   &\text{where} \quad s =  S(\beta) = 2 \sinh^2(\beta/2).
\end{aligned}
\end{equation}
%
The ``sticky'' models are optimal at values of $s$ smaller than this.
This is not quite the same as the lower boundary of the region \eqref{eq:univalid}, so there is a gap between them where the uniform model with $M$ states is optimal.
The width of this gap is insignificant for larger $M$.

@@

To find the optimal \(\varepsilon\) at any given \(s\), it helps if we define
%
\begin{equation}\label{eq:shortdefs}
\begin{aligned}
  \Delta f_m(\beta) &= S(m\beta) - \zeta S((m-1)\beta), &
  f_m(\beta) &= \frac{S(m\beta)}{\Delta f_m(\beta)}, \\
  \Delta g_m(\beta) &= \Delta f_m(\beta) + 1 - \zeta , &
  g_m(\beta) &= \frac{S(m\beta) + 1}{\Delta g_m(\beta)}, \\
  \text{where}\quad \zeta &= 2s + 1. \\
\end{aligned}
\end{equation}
%
Then \cref{eq:shortenlaplace} can be be written as
%
\begin{equation}\label{eq:shortsimple}
  \hat{A}(s) =
    = \brk{\frac{\Delta f_{m}}{s \Delta g_{m}}}
    \frac{f_m - \varepsilon}{(m - \varepsilon) (g_m - \varepsilon)}.
\end{equation}
%

We can maximise this \wrt $\varepsilon$.
%
\begin{equation}\label{eq:shortopt}
  (\varepsilon - f_m)^2 = (m - f_m)(g_m - f_m),
  \qquad
  \hat{A}(s) = \brk{\frac{\Delta f_{m}}{s \Delta g_{m}}}
    \frac{1}{\brk{\sqrt{\abs{m - f_m}} - \sqrt{\abs{g_m - f_m}}}^2}.
\end{equation}
%
We can find the boundary of the region of applicability by asking at what value of $s$ is the optimal value of $\varepsilon=0$ or $1$?
%
\begin{equation}\label{eq:shortenvalid}
\begin{aligned}
  \left.\pdiff{\hat{A}(s)}{\varepsilon}\right|_{\varepsilon=0} \propto &\,
    m(2s+1) \sqrt{2s} \sinh\brk{{\prn{m-\tfrac{1}{2}}\beta}} 
    - S(m\beta) [S(m\beta) + 1]
      = 0,\\
  \left.\pdiff{\hat{A}(s)}{\varepsilon}\right|_{\varepsilon=1} \propto &\,
    (m-1) \sqrt{2s} \sinh\brk{{\prn{m-\tfrac{1}{2}}\beta}} 
    - (2s+1) S((m-1)\beta) [S((m-1)\beta) + 1]
      = 0,\\
    &\text{where}\quad
    s = S(\beta) = 2 \sinh^2(\beta/2),
\end{aligned}
\end{equation}
%
with the $\varepsilon=0$ case providing the lower boundary and $\varepsilon=1$ providing the upper boundary.
For lower \(s\), \(\varepsilon = 0\) is the optimum. 
For higher \(s\), \(\varepsilon = 1\) is the optimum, which is eqivalent to \(\varepsilon = 0\) with \(M-2\) states.
This is depicted by the dotted lines in \cref{fig:shorten}\ref{fig:short_ratio}.

@@

We will phrase it in terms of $\snrb(\tau)$, reintroducing $r$:
%
\begin{equation}\label{eq:envunits}
  \snrb(\tau) = \frac{ \sqrt{N} \hat{A}\prn{\frac{1}{r\tau}} }{ r\tau },
  \qquad
  \alpha(\tau) = \frac{ 1 + r\tau +\sqrt{1+2r\tau} }{ r\tau }.
\end{equation}
%
Then, combining \eqref{eq:binaryenv}, \eqref{eq:unienv} and \eqref{eq:stickylaplace}, we find
%
\begin{equation}\label{eq:henv}
  \snrb(\tau) = \left\{
\begin{aligned}
  & \frac{\sqrt{N}}{1+r\tau}, &
  & r\tau \leq \tfrac{1}{2\sinh^2\frac{y_*}{2}} , \\
  & \sqrt{N}\prn{ \frac{ 4\sinh^2 \frac{y_*}{2} }{ y_* \cosh y_* } }\sinh\inv \!\!\sqrt{1/2r\tau}, &
  \tfrac{1}{2\sinh^2 \frac{y_*}{2}} \leq & r\tau \leq \tfrac{1}{2\sinh^2 \frac{y_*}{M}} , \\
  & \max_{\varepsilon \in [0,1]} \brc{
      \frac{2\sqrt{N}(1-\varepsilon) \Delta_{M,\varepsilon} \brk{S\prn{\tfrac{M\beta}{2}}}}
      {\Delta_{M,\varepsilon}[M]\, \Delta_{M,\varepsilon}\! \brk{S\prn{\tfrac{M\beta}{2}} + 1} } }, &
  \tfrac{1}{2\sinh^2 \frac{y_*}{M}} \leq & r\tau ,
\end{aligned}
\right.
\end{equation}
%
where  we use the notation \( \Delta_{M,\varepsilon}[T(M, \beta)] = T(M, \beta) - \varepsilon T(M-2, \beta) \).
For $M \gg 1$, and $r\tau$ much greater than the lower bound of each region, we can approximate this as
%
\begin{equation}\label{eq:henvapproc}
  \snrb(\tau) \approx \left\{
\begin{aligned}
  & \frac{\sqrt{N}}{1+r\tau}, &
  & r\tau \leq 0.73 , \\
  & \frac{ 0.54\sqrt{N}}{ \sqrt{r\tau} } , &
  0.73 \leq & r\tau \leq 0.22 M^2 , \\
  & \frac{ \sqrt{N} (M-1) }{ r\tau }, &\quad
  0.22 M^2 \leq &r\tau .
\end{aligned}
\right.
\end{equation}
%
This is plotted in \autoref{fig:heuristicenv}.

>>>>

% -*- TeX -*- -*- UK -*-
% ----------------------------------------------------------------
% arXiv Paper ************************************************
%
% Subhaneil Lahiri's template
%
% Before submitting:
%    Comment out hyperref
%    Comment out showkeys
%    Replace \input{?.tex} with its contents
%       or include ?.tex in zip/tar file
%    Put this file, the .bbl file, any picture or
%       other additional files and natbib.sty
%       file in a zip/tar file
%
% **** -----------------------------------------------------------
\documentclass[12pt]{article}
%Preamble:
\usepackage{a4wide}
\input{sl_preamble.tex}
\usepackage{mathtools}
%
% >> Only for drafts! <<
\usepackage[notref,notcite]{showkeys}
%
\input{sl_graphics_preamble.tex}
\graphicspath{{Figures/}}
%
% ----------------------------------------------------------------
%\numberwithin{equation}{section}
%\renewcommand{\baselinestretch}{1.5}
% ----------------------------------------------------------------
% New commands etc.
\input{sl_definitions.tex}
\input{sl_symbols.tex}
%matrices
%vec of ones
\newcommand{\onev}{\mathbf{e}}
%mat of ones
\newcommand{\onem}{\mathbf{E}}
%Markov matrix
\newcommand{\MM}{\mathbf{Q}}
%prob distribution
\newcommand{\pr}{\mathbf{p}}
%equilibrium distribution
\newcommand{\eq}{\pr^\infty}
%first passage times
\newcommand{\fpt}{\mathbf{T}}
%off-diag first passage times
\newcommand{\fptb}{\overline{\fpt}}
%fundamental matrix
\newcommand{\fund}{\mathbf{Z}}
%other symbols
\newcommand{\Pb}{\mathbf{P}}
\newcommand{\D}{\mathbf{D}}
\newcommand{\pib}{\boldsymbol{\pi}}
\newcommand{\Lb}{\boldsymbol{\Lambda}}
\newcommand{\etw}{\boldsymbol{\eta}^w}
\newcommand{\thb}{\boldsymbol{\theta}}
%synapse models
\newcommand{\w}{\mathbf{w}}
\newcommand{\W}{\mathbf{W}}
\newcommand{\M}{\mathbf{M}}
\newcommand{\enc}{\mathbf{K}}
\newcommand{\frg}{\W^{\mathrm{F}}}
\newcommand{\F}{\boldsymbol{\Phi}}
%superscripts
\newcommand{\pot}{^{\text{pot}}}
\newcommand{\dep}{^{\text{dep}}}
\newcommand{\potdep}{^{\text{pot/dep}}}
%with activity indep processes
\renewcommand{\hom}{\mathbf{H}}
\newcommand{\Mh}{\widetilde{\M}}
\newcommand{\frgh}{\widetilde{\W}^{\mathrm{F}}}
\newcommand{\ench}{\widetilde{\enc}}
%eigenvectors
\newcommand{\evr}{\mathbf{u}}
\newcommand{\evl}{\boldsymbol{\eta}}
%snr curves etc
\newcommand{\syn}{\vec{w}}
\newcommand{\synid}{\syn_\text{id}}
\DeclareMathOperator{\SNR}{SNR}
\DeclareMathOperator{\snr}{SNR}
\DeclareMathOperator{\NNR}{NNR}
\DeclareMathOperator{\nnr}{NNR}
\newcommand{\snrb}{\overline{\snr}}
\DeclareMathOperator{\env}{Env}
\newcommand{\rh}{\hat{r}}
\newcommand{\CA}{\mathcal{A}}
\newcommand{\CB}{\mathcal{B}}
%avoid double superscripts
\renewcommand{\pdiff}[2]{\frac{\partial #1}{\partial #2}}

@@

In reality, $M$ can only take even integer values.
However, one can find a heuristic approximation to the envelope by maximising \wrt $M$.
We find that
%
\begin{multline}\label{eq:unienv}
  M_* = \pm\frac{2y_*}{\beta},
  \quad
  s = 2\sinh^2\frac{y_*}{M_*},
  \qquad
  \hat{A}_*(s) = \prn{\frac{ 2\sinh^2 \frac{y_*}{2} }{ y_* \cosh y_* } } \frac{\abs{\beta}}{s},
  \\ \text{where}\quad
  y_* = \tanh \frac{y_*}{2} \cosh y_* \approx 1.50553.
  \qquad
\end{multline}
%
This is only valid for
%
\begin{equation}\label{eq:univalid}
  2 \leq M_* \leq M,
  \qquad\implies\qquad
  2\sinh^2 \frac{y_*}{M} \leq s \leq 2\sinh^2 \frac{y_*}{2}.
\end{equation}
%
When $M \gg 2$, the validity region is approximately
%
\begin{equation}\label{eq:univalidapprox}
  \frac{4.53327}{M^2} \lesssim s \lesssim 1.36423.
\end{equation}
%
When $s \sim \CO(M^{-2}) \ll 1$, we have \(\beta \approx \pm \sqrt{2s}\) and the heuristic envelope is approximately
%
\begin{equation}\label{eq:unienvapprox}
  \hat{A}_*(s) \approx \frac{ 0.54203 }{ \sqrt{s} }.
\end{equation}
%
We will lok at a more accurate interpolation between different \(M\) in the \hyperref[sec:serial_shorten]{next section}.

@@

To find the optimal \(\varepsilon\) at any given \(s\), it helps if we define
%
\begin{equation*}
\begin{aligned}
  \Delta f_m(\beta) &= S(m\beta) - \zeta S((m-1)\beta), &
  f_m(\beta) &= \frac{S(m\beta)}{\Delta f_m(\beta)}, \\
  \Delta g_m(\beta) &= \Delta f_m(\beta) + 1 - \zeta , &
  g_m(\beta) &= \frac{S(m\beta) + 1}{\Delta g_m(\beta)}, \\
  \text{where}\quad \zeta &= 2s + 1, 
  \qquad\qquad \implies & 
  \hat{A}(s) &= \frac{\Delta f_{m} (f_m - \varepsilon)}
           {s \Delta g_{m} (m - \varepsilon) (g_m - \varepsilon)}.
  \\
\end{aligned}
\end{equation*}
%
The results are plotted in \cref{fig:shorten}.

We can maximise this \wrt $\varepsilon$.
%
\begin{equation}\label{eq:shortopt}
  (\varepsilon - f_m)^2 = (m - f_m)(g_m - f_m),
  \qquad
  \hat{A}(s) = \frac{\Delta f_{m}}
      {s \Delta g_{m} \brk{\sqrt{\abs{m - f_m}} - \sqrt{\abs{g_m - f_m}}}^2}.
\end{equation}
%
We can find the boundary of the region of applicability by asking at what value of $s$ is the optimal value of $\varepsilon=0$ or $1$?
%
\begin{equation}\label{eq:shortenvalid}
\begin{aligned}
  \left.\pdiff{\hat{A}(s)}{\varepsilon}\right|_{\mathrlap{\varepsilon=0}} \propto &\,
    m(2s+1) \brk{S(m\beta) - S((m-1)\beta)} - S(m\beta) [S(m\beta) + 1]
      = 0,\\
  \left.\pdiff{\hat{A}(s)}{\varepsilon}\right|_{\mathrlap{\varepsilon=1}} \propto &\,
    (m-1)\brk{S(m\beta) - S((m-1)\beta)} - (2s+1) S((m-1)\beta) [S((m-1)\beta) + 1]
      = 0,\\
    &\text{where}\quad
    s = S(\beta) = 2 \sinh^2(\beta/2),
\end{aligned}
\end{equation}
%
with the $\varepsilon=0$ case providing the lower boundary and $\varepsilon=1$ providing the upper boundary.
For lower \(s\), \(\varepsilon = 0\) is the optimum. 
For higher \(s\), \(\varepsilon = 1\) is the optimum, which is eqivalent to \(\varepsilon = 0\) with \(M-2\) states.
This is depicted by the dotted lines in \cref{fig:shorten}\ref{fig:short_ratio}.

@@

\begin{figure}[ht]
\begin{center}
  \includegraphics[width=0.4\linewidth]{sticky_opt.pdf}
\caption{Signal-to-noise ratio for optimal sticky models, \cref{eq:stickyenc} optimised over \(\varepsilon\) for \(M=10\), in blue.
  The orange line is the SNR for the optimised shortened models, \cref{eq:shortenenc} with \(M=10\).
  The grey line is our theoretical envelope from \cref{eq:env}.
  \label{fig:sticky}}
\end{center}
\end{figure}

We can maximise this \wrt $\varepsilon$.
This results in a complicated quadratic equation for $\varepsilon$ with an even worse solution that we will not reproduce here.
We define
%
\begin{equation*}
\begin{aligned}
  \Delta m &= (m-1), &
  \mu &= \frac{m}{\Delta m}, \\
  \Delta f_m(\beta) &= S((m-1)\beta), &
  f_m(\beta) &= \frac{S(m\beta)}{\Delta f_m(\beta)}, \\
  \Delta g_m(\beta) &= \Delta f_m(\beta) + 1 , &
  g_m(\beta) &= \frac{S(m\beta) + 1}{\Delta g_m(\beta)}, \\
\end{aligned}
 \; \implies \quad
 \hat{A}(s) = \frac{\Delta f_m (1 - \varepsilon) (f_m - \varepsilon)}
                    {s \Delta m \Delta g_m (\mu - \varepsilon) (g_m - \varepsilon)}.
\end{equation*}
%
The minimum is at the solution of
%
\begin{equation*}
  \brk{(1 + f_m -  g_m - \mu)\varepsilon - f_m + \mu g_m}^2 
    = (\mu - 1) (g_m - 1) (\mu - f_m) (g_m - f_m).
\end{equation*}
%
The result is plotted in \cref{fig:sticky}.

We can find the boundary of the region of applicability by seeing at what value of $s$ is the optimal value of $\varepsilon=0$:
%
\begin{equation}\label{eq:stickyvalid}
\begin{aligned}
  \left.\pdiff{\hat{A}(s)}{\varepsilon}\right|_{\mathrlap{\varepsilon=0}} &\propto
    m \brk{S(m\beta) - S((m-1)\beta)} - S(m\beta) \brk{S(m\beta) + 1} = 0,\\
   &\text{where} \quad s =  S(\beta) = 2 \sinh^2(\beta/2).
\end{aligned}
\end{equation}
%
The ``sticky'' models are optimal at values of $s$ smaller than this.
This is not quite the same as the lower boundary of the region \cref{eq:shortenvalid}, or its approximate version \cref{eq:univalid}, so there is a gap between them where the uniform model with $M$ states is optimal.
The width of this gap is insignificant for larger $M$, as seen in \cref{fig:sticky}.

@@

@@

We will phrase it in terms of $\snrb(\tau)$, reintroducing $r$:
%
\begin{equation}\label{eq:envunits}
  \snrb(\tau) = \frac{ \sqrt{N} \hat{A}\prn{\frac{1}{r\tau}} }{ r\tau },
  \qquad
  \beta(\tau) = \sinh\inv \!\!\sqrt{1/2r\tau}.
\end{equation}
%
Then, combining \eqref{eq:binaryenv}, \eqref{eq:unienv} and \eqref{eq:stickylaplace}, using the notation \( \Delta_{M,\varepsilon}[T(M, \beta)] = T(M, \beta) - \varepsilon T(M-2, \beta) \)., we find
%
\begin{equation}\label{eq:henv}
  \snrb(\tau) = \left\{
\begin{aligned}
  & \frac{\sqrt{N}}{1+r\tau}, &
  & r\tau \leq \tfrac{1}{2\sinh^2\frac{y_*}{2}} , \\
  & \sqrt{N}\prn{ \frac{ 4\sinh^2 \frac{y_*}{2} }{ y_* \cosh y_* } } \beta(\tau), &
  \tfrac{1}{2\sinh^2 \frac{y_*}{2}} \leq & r\tau \leq \tfrac{1}{2\sinh^2 \frac{y_*}{M}} , \\
  & \max_{\varepsilon \in [0,1]} \brc{
      \frac{2\sqrt{N}(1-\varepsilon)\, \Delta_{M,\varepsilon}\! \brk{S\prn{\tfrac{M\beta}{2}}}}
      {\Delta_{M,\varepsilon}[M]\, \Delta_{M,\varepsilon}\! \brk{S\prn{\tfrac{M\beta}{2}} + 1} } }, &
  \tfrac{1}{2\sinh^2 \frac{y_*}{M}} \leq & r\tau .
\end{aligned}
\right.
\end{equation}
%
For $M \gg 1$, and $r\tau$ much greater than the lower bound of each region, we can approximate this as
%
\begin{equation}\label{eq:henvapproc}
  \snrb(\tau) \approx \left\{
\begin{aligned}
  & \frac{\sqrt{N}}{1+r\tau}, &
  & r\tau \leq 0.73 , \\
  & \frac{ 0.54\sqrt{N}}{ \sqrt{r\tau} } , &
  0.73 \leq & r\tau \leq 0.22 M^2 , \\
  & \frac{ \sqrt{N} (M-1) }{ r\tau }, &\quad
  0.22 M^2 \leq &r\tau .
\end{aligned}
\right.
\end{equation}
%
This is plotted in \autoref{fig:heuristicenv}.
